Missing the Point: Non-Convergence in Iterative Imputation Algorithms

### 2020-05-06 3.Thesis/Writeup/2.Manuscript.Rmd
Reread Results until thetas

### $Q=\beta$.
For the estimated regression coefficients, we consider bias, confidence interval width (CIW), and coverage rate (CR) as performance measures.  Unbiased estimates are obtained in conditions where $p_{mis}\leq.50$ and $T\geq3$. Approximately confidence valid estimates are obtained in any condition where $T\geq3$.

Conditions with a lower number of iterations yield more bias in $\bar{Q}$. This is especially pronounced in conditions with high missingness proportions: with 95% missingness we need at least seven iterations before an additional iteration does not lead to improved estimates. Conditions with a lower missingness proportion (e.g., $p_{mis} = {0.5}$), produce *optimal* estimates with as many as two iterations. Approximately unbiased estimates are obtained for all conditions where $p_{mis}\leq0.75$ and $T\leq6$. **[Add CIW and CR]**

@@

In Figure \ref{fig:rhats-acs} we see the results of the non-convergence diagnostics $\widehat{R}$ and $AC$. **[Add panel labels and explain how to read the plots]** As expected, the diagnostics identify more signs of non-convergence in simulation conditions with a lower number of iterations. The missingness proportions, however, only seem to impact multivariate convergence. When $\widehat{R}$ and $AC$ are applied to  univariate $\theta$s, they identify the same amount of non-convergence, irrespective of the missingness proportion. When the diagnostics are applied to multivariate $\theta$s, they generally identify more signs of non-convergence for higher missingness proportions.

None of the simulation conditions that we consider resulted in complete convergence of the algorithm (defined as $\widehat{R}=1$ and $AC=0$). The thresholds for diagnosing non-convergence that are used in practice seem too conservative compared to the performance measures: $\widehat{R}$ and $AC$ identify signs of non-convergence in conditions where the $\bar{Q}$s are unbiased and confidence valid estimates of the $Q$s. The most recent recommended threshold for diagnosing non-convergence with $\widehat{R}$ is never overcome in this study (all $\widehat{R}$>1.01). Apparently, this threshold is too stringent for iterative imputation algorithms, and will not be discussed further.

>>>>

### $Q=\beta$.
For the estimated regression coefficients, we consider bias, confidence interval width (CIW), and coverage rate (CR) as performance measures. Unbiased estimates are obtained in conditions where $p_{mis}\leq.50$ and $T\geq3$. Approximately confidence valid estimates are obtained in any condition where $T\geq3$---despite of persistant bias in conditions where $p_{mis}\geq.75$. These approximately nominal coverages are explained by the CIWs: conditions with higher missingness proportions result in more uncertainty in the estimates (i.e., wider CIs), which diminish the effect of the bias. *Sufficient* inferences in terms of confidence validity may be obtained when $T\geq3$, while we need at least seven iterations to reach this point in terms of bias. Continued iteration after $T\geq7$ seems futile.
<!-- One exception is the CIW of the most extreme condition ($p_{mis}=.95$; $T=1$). -->

<!-- Conditions with a lower number of iterations yield more bias in $\bar{Q}$. This is especially pronounced in conditions with high missingness proportions: with 95% missingness we need at least seven iterations before an additional iteration does not lead to improved estimates. Conditions with a lower missingness proportion (e.g., $p_{mis} = {0.5}$), produce *optimal* estimates with as many as two iterations. Approximately unbiased estimates are obtained for all conditions where $p_{mis}\leq0.75$ and $T\leq6$. **[Add CIW and CR]** -->

@@

We evaluate the non-convergence diagnostics $\widehat{R}$ and $AC$ against the performance measures described above (i.e., bias and confidence validity of the quantities of scientific interest $Q$). The diagnostics should identify those simulation conditions that are negatively impacted by non-convergence. We differentiate between approximately unbiased and confidence valid estimates, and *sufficiently* valid estimates. The former is defined by the interplay between $Q$, $p_{mis}$, and $T$. The latter is observed in all simulation conditions where $T\geq7$, irrespective of the $Q$ and $p_{mis}$. We thus expect the most signs of non-convergence in conditions where $T<7$.

<!-- As we have observed, the affected conditions are $p_{mis}=.95$ for all all $Q$s are impacted by the missingness proportions, but only the multivariate quantities of scientific interest are affected by the number of iterations.  -->

Figure \ref{fig:rhats-acs} displays the non-convergence diagnostics $\widehat{R}$ and $AC$ for each of the four $\theta$s under consideration. **[Add panel labels and explain how to read the plots]** In general, the diagnostics identify more signs of non-convergence in simulation conditions with a lower number of iterations. The missingness proportions, however, only seem to impact multivariate convergence. When $\widehat{R}$ and $AC$ are applied to  univariate $\theta$s, they identify the same amount of non-convergence for any missingness proportion. When the diagnostics are applied to multivariate $\theta$s, they generally identify more signs of non-convergence for higher missingness proportions.

As expected, none of the simulation conditions that we consider resulted in complete convergence of the algorithm (defined as $\widehat{R}=1$ and $AC=0$). Yet, the thresholds used in practice to diagnose non-convergence seem too conservative as well.  $\widehat{R}$ and $AC$ identify signs of non-convergence in conditions where the $\bar{Q}$s are unbiased and confidence valid estimates of the $Q$s. The most recent recommended threshold for diagnosing non-convergence with $\widehat{R}$ is never overcome in this study (all $\widehat{R}$>1.01). Apparently, this threshold is too stringent for iterative imputation algorithms, and will not be discussed any further.

### 2020-05-06 3.Thesis/Writeup/2.Manuscript.tex
Reread Results until thetas

Figure \ref{fig:Qs} displays the influence of inducing non-convergence on the estimated quantities of scientific interest, \(Q\)s. \textbf{{[}Add panel labels, describe panels{]}} As expected, conditions with a higher proportion of missingness and/or a lower number of iterations are impacted more severely by non-convergence. I.e., on average these conditions portray more extreme bias in the estimated \(Q\)s, and non-nominal coverages. Roughly speaking, this means that the algorithm is indeed not converged at \(t=1\), and converges gradually as \(t\) increases to \(T\). The point at which an additional iteration does not lead to an improvement in the estimates depends on quantity of scientific interest \(Q\) and the missingness proportion \(p_{mis}\). The highest number of iterations that is necessary to reach this point across all simulation conditions is \(T=7\). This implies that, under the current specifications, no more than seven iterations are necessary to obtain \emph{sufficient} performance in terms of bias and confidence validity.

@@

The estimated univariate means seem unaffected by the number of iterations in the algorithm. The magnitude of the bias in \(\bar{Q}\) depends exclusively on the missingness proportion \(p_{mis}\), with higher missingness proportions causing more extreme bias. Note however, that the simulation condition with the largest overall bias resulted in a \(\bar{Q}\) under-estimated the true value of \(Q\) by only 0.02 units (where \(Q=\mu_Y=25.81; \sigma_Y = 11.32\)). We conclude that---conditional on the missingness proportion---all simulation conditions (\(T \geq 1\)) have equal performance. According to this scientific estimand \(Q\), non-convergence does not impact the validity of the inferences. \textbf{{[}Add: up-to 50\% missingness everything is relatively unbiased, while 75\% and 95\% missingness show very different results. Why is that? Add sd of Y in the figure to show that the bias is relative: a bias of 0.02 for a variable with a mean of 25.81, and sd of 11.32 is nothing.{]}}

@@

The estimated standard deviations show impact of non-convergence similar to the means. The magnitude of the bias \(\bar{Q}\) follows the incremental order of the missingness proportions. The number of iterations in the algorithm has little effect on the bias. Only the estimates in simulation conditions with just one iteration (\(T=1\)) differ from other conditions that have the same missingness proportion but a higher number of iterations. Interestingly, the bias in conditions where \(T=1\) is \emph{less} severe than after continued iteration. For the least possible bias, we may use any number of iterations \(T\geq1\), but only \(p_{mis} = 0.05\) results in completely unbiased \(\bar{Q}\)s.

@@

For the estimated regression coefficients, we consider bias, confidence interval width (CIW), and coverage rate (CR) as performance measures. The magnitude of the bias, again, depends on the missingness proportion. But for this \(Q\) we also observe a clear effect of the number of iterations. Conditions with a lower number of iterations yield more bias in \(\bar{Q}\). This is especially pronounced in conditions with high missingness proportions: with 95\% missingness we need at least seven iterations before an additional iteration does not lead to improved estimates. Conditions with a lower missingness proportion (e.g., \(p_{mis} = {0.5}\)), produce \emph{optimal} estimates with as many as two iterations. Approximately unbiased estimates are obtained for all conditions where \(p_{mis}\leq0.75\) and \(T\leq6\). \textbf{{[}Add CIW and CR{]}}

The bias in the estimated explained variance perfectly follows both sets of simulation conditions: The worst performance is observed for the highest missingness proportion and the lowest number of iterations, while the best performance occurs with less missingness and more iterations. Only conditions with the lowest missingness proportion (\(p_{mis}=0.05\)) results in unbiased estimates for every number of iterations. In other conditions there is at least some bias irrespective of \(T\). The inferences do not improve in any condition where \(T \geq 6\). This suggests that the algorithm is as converged as necessary to obtain valid inferences.

@@

In Figure \ref{fig:rhats-acs} we see the results of the non-convergence diagnostics \(\widehat{R}\) and \(AC\). \textbf{{[}Add panel labels and explain how to read the plots{]}} As expected, the diagnostics identify more signs of non-convergence in simulation conditions with a lower number of iterations. The missingness proportions, however, only seem to impact multivariate convergence. When \(\widehat{R}\) and \(AC\) are applied to univariate \(\theta\)s, they identify the same amount of non-convergence, irrespective of the missingness proportion. When the diagnostics are applied to multivariate \(\theta\)s, they generally identify more signs of non-convergence for higher missingness proportions.

None of the simulation conditions that we consider resulted in complete convergence of the algorithm (defined as \(\widehat{R}=1\) and \(AC=0\)). The thresholds for diagnosing non-convergence that are used in practice seem too conservative compared to the performance measures: \(\widehat{R}\) and \(AC\) identify signs of non-convergence in conditions where the \(\bar{Q}\)s are unbiased and confidence valid estimates of the \(Q\)s. The most recent recommended threshold for diagnosing non-convergence with \(\widehat{R}\) is never overcome in this study (all \(\widehat{R}\)\textgreater1.01). Apparently, this threshold is too stringent for iterative imputation algorithms, and will not be discussed further.

>>>>

Figure \ref{fig:Qs} displays the influence of inducing non-convergence on the estimated quantities of scientific interest, \(Q\)s. \textbf{{[}Add panel labels, describe panels{]}} Overall, conditions with a higher proportion of missingness and/or a lower number of iterations are impacted more severely by non-convergence. I.e., on average these conditions portray more extreme bias in the estimated \(Q\)s, and non-nominal coverages. Roughly speaking, this means that the algorithm is indeed not converged at \(t=1\), and converges gradually as \(t\) increases to \(T\). The point at which an additional iteration does not lead to an improvement in the estimates depends on quantity of scientific interest \(Q\) and the missingness proportion \(p_{mis}\). The highest number of iterations that is necessary to reach this point across all simulation conditions is \(T=7\). This implies that, under the current specifications, no more than seven iterations are necessary to obtain \emph{sufficient} performance in terms of bias and confidence validity.

@@

The estimated univariate means seem unaffected by the number of iterations in the algorithm. This implies that \emph{suffient} inferences may be obtained with as little as one iteration (\(T \geq 1\)). The magnitude of the bias in \(\bar{Q}\) depends exclusively on the missingness proportion \(p_{mis}\), with missingness proportions of 75\% and 95\% causing more extreme bias than other conditions. Unbiased estimates are obtained in all conditions where \(p_{mis}\leq.50\). Note however, that even in the conditions where \(p_{mis}\geq.75\), the bias is relatively small: \(\mu_Y\) is maximally under-estimated by 0.02 units, while the true value of \(\mu_Y\) is 25.81 (\(\sigma_Y = 11.32\)). We conclude that non-convergence does not substantially impact the validity of the inferences for \(Q=\mu\). \textbf{{[}Add: up-to 50\% missingness everything is relatively unbiased, while 75\% and 95\% missingness show very different results. Why is that? Add sd of Y in the figure to show that the bias is relative: a bias of 0.02 for a variable with a mean of 25.81, and sd of 11.32 is nothing.{]}}

@@

The estimated standard deviations are largely affected by the missingness proportions, but not to much by the number of iterations. \emph{Sufficient} inferences are obtained in conditions with just two iterations. Interestingly, the bias in first iteration is actually \emph{less} extreme than for any other number of iterations. We may therefore conclude that no iteration at all is needed to obtain \emph{sufficient} estimates for \(Q=\sigma\). Completely unbiased estimates are only obtained in conditions where \(p_{mis}=.05\). Conditions where \(p_{mis}\geq.25\) are impacted by non-convergence in order of increasing missingness proportion.

Bias in the estimated variance explained depends both on missingness proportions and the number of iterations. The magnitude of the bias clearly follows the incremental order of the missingness proportions. We, therefore, observe \emph{sufficient} estimates at a different number of iterations for each missingness proportion. For \(p_{mis}\leq.25\), the minimum number of iterations is two (\(T\geq2\)), while \(p_{mis}=.50\) requires \(T\geq4\), and \(p_{mis}=.75\) requires \(T\geq5\). The highest number of iterations needed to obtain \emph{sufficient} estimates is seven (\(p_{mis}=.95\); \(T\geq7\)). Unbiased estimates, however, may only be obtained if the missingness proportion is at most 25\%. \textbf{{[}interpret sign of bias{]}}

@@

For the estimated regression coefficients, we consider bias, confidence interval width (CIW), and coverage rate (CR) as performance measures. Unbiased estimates are obtained in conditions where \(p_{mis}\leq.50\) and \(T\geq3\). Approximately confidence valid estimates are obtained in any condition where \(T\geq3\)---despite of persistant bias in conditions where \(p_{mis}\geq.75\). These approximately nominal coverages are explained by the CIWs: conditions with higher missingness proportions result in more uncertainty in the estimates (i.e., wider CIs), which diminish the effect of the bias. \emph{Sufficient} inferences in terms of confidence validity may be obtained when \(T\geq3\), while we need at least seven iterations to reach this point in terms of bias. Continued iteration after \(T\geq7\) seems futile.

@@

We evaluate the non-convergence diagnostics \(\widehat{R}\) and \(AC\) against the performance measures described above (i.e., bias and confidence validity of the quantities of scientific interest \(Q\)). The diagnostics should identify those simulation conditions that are negatively impacted by non-convergence. We differentiate between approximately unbiased and confidence valid estimates, and \emph{sufficiently} valid estimates. The former is defined by the interplay between \(Q\), \(p_{mis}\), and \(T\). The latter is observed in all simulation conditions where \(T\geq7\), irrespective of the \(Q\) and \(p_{mis}\). We thus expect the most signs of non-convergence in conditions where \(T<7\).

Figure \ref{fig:rhats-acs} displays the non-convergence diagnostics \(\widehat{R}\) and \(AC\) for each of the four \(\theta\)s under consideration. \textbf{{[}Add panel labels and explain how to read the plots{]}} In general, the diagnostics identify more signs of non-convergence in simulation conditions with a lower number of iterations. The missingness proportions, however, only seem to impact multivariate convergence. When \(\widehat{R}\) and \(AC\) are applied to univariate \(\theta\)s, they identify the same amount of non-convergence for any missingness proportion. When the diagnostics are applied to multivariate \(\theta\)s, they generally identify more signs of non-convergence for higher missingness proportions.

As expected, none of the simulation conditions that we consider resulted in complete convergence of the algorithm (defined as \(\widehat{R}=1\) and \(AC=0\)). Yet, the thresholds used in practice to diagnose non-convergence seem too conservative as well. \(\widehat{R}\) and \(AC\) identify signs of non-convergence in conditions where the \(\bar{Q}\)s are unbiased and confidence valid estimates of the \(Q\)s. The most recent recommended threshold for diagnosing non-convergence with \(\widehat{R}\) is never overcome in this study (all \(\widehat{R}\)\textgreater1.01). Apparently, this threshold is too stringent for iterative imputation algorithms, and will not be discussed any further.
